<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.path = 'man/figures/README-',
  out.width = '75%',
  dpi = 110,
  message = FALSE,
  warning = FALSE
)
options(width = 90, tibble.print_min = 8, tibble.print_max = 20)
set.seed(1)
devtools::load_all('.', quiet = TRUE)
suppressWarnings(suppressMessages(
  set_plot_font(font = 'sans', size = 12, search_sources = 'system')
))
```

# kkstatfun

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.18936020.svg)](https://doi.org/10.5281/zenodo.18936020)

**R Statistical Analysis Toolkit for Medical and Epidemiology**

A comprehensive R package designed for medical statistics, epidemiology, and general clinical data analysis. It provides a suite of tidyverse-friendly tools with full support for piping (`%>%` / `|>`) and grouping (`group_by()`).

---

## Installation

You can install the package directly from GitHub:

```{r ex-01, eval=FALSE}
# Install devtools if not already installed
if (!require("devtools")) install.packages("devtools")

# Install kkstatfun from GitHub
devtools::install_github("kostadinoff/kkstatfun")
```

---

## Comprehensive Function Reference (with Epi & Clinical Examples)

### 1. Epidemiology & Measures of Association

#### `kk_twobytwo(data, exposure, outcome)`

Computes classic measures of association from $2 \times 2$ contingency tables: Odds Ratio (OR), Relative Risk (RR), Risk Difference (RD), Attributable Fraction (AF), Population Attributable Fraction (PAF), and Preventable Fraction (PF) with 95% confidence intervals.

*   **Epidemiological Example**: Assessing the risk of developing lung cancer among smokers vs. non-smokers.

```{r ex-02}
library(kkstatfun)
df <- data.frame(
  smoking = c(rep(1, 80), rep(0, 120)),
  lung_cancer = c(rep(1, 30), rep(0, 50), rep(1, 10), rep(0, 110))
)
df %>% kk_twobytwo(smoking, lung_cancer)
```

> **Interpretation.** Smokers had 4.5× the risk of lung cancer (RR 4.5, 95% CI 2.3–8.7) and 6.6× the odds (OR 6.6). Because the outcome is common (20% overall), the OR overstates the RR — report the RR here. The risk difference of 0.29 means ~29 extra cases per 100 smokers, i.e. one extra case for every ~3.4 smokers (NNH 3.4). 78% of cancers *among smokers* are attributable to smoking (AF), and 58% of *all* cancers in the population would be prevented if smoking were removed (PAF). Every CI excludes the null, so the association is significant.

#### `kk_stratified_2x2(data, exposure, outcome, stratum)`

Performs stratified analysis on $2 \times 2$ tables, calculating stratum-specific ORs, the Cochran-Mantel-Haenszel (CMH) pooled Odds Ratio, and the Breslow-Day test for homogeneity of odds ratios to detect confounding and effect modification.

*   **Epidemiological Example**: Studying the association between coffee consumption and coronary heart disease, stratified by smoking status to control for confounding.
```{r ex-03}
df_strat <- data.frame(
  coffee = c(rep(1, 50), rep(0, 50), rep(1, 40), rep(0, 60)),
  chd = c(rep(1, 15), rep(0, 35), rep(1, 5), rep(0, 45), rep(1, 10), rep(0, 30), rep(1, 12), rep(0, 48)),
  smoking = c(rep("Smoker", 100), rep("Non-Smoker", 100))
)
df_strat %>% kk_stratified_2x2(coffee, chd, smoking)
```

> **Interpretation.** Pooling across smoking strata, coffee is associated with 2.16× the odds of CHD (Mantel-Haenszel OR 2.16, 95% CI 1.07–4.35). The Breslow-Day test is non-significant (p = 0.085), so the odds ratios are reasonably *homogeneous* across strata — no strong evidence of effect modification, and the single pooled estimate is a fair summary. The stratum-specific ORs (3.86 in smokers vs 1.33 in non-smokers) hint at a stronger effect among smokers, but the sample is too small to confirm interaction.

#### `kk_reri(data, exp1, exp2, outcome)`

Calculates the Relative Excess Risk due to Interaction (RERI), Attributable Proportion due to Interaction (AP), and Synergy Index (S) to evaluate additive interactions.

*   **Epidemiological Example**: Assessing if the combined effect of asbestos exposure and smoking on lung cancer is greater than the sum of their individual effects.
```{r ex-04}
df_int <- data.frame(
  asbestos = rep(c(1, 0), each = 100),
  smoking = rep(c(1, 0, 1, 0), each = 50),
  cancer = rbinom(200, 1, 0.3)
)
df_int %>% kk_reri(asbestos, smoking, cancer)
```

> **Interpretation.** RERI is the excess risk on the *additive* scale beyond the sum of the two separate exposure effects. Here RERI = 0.44 but its 95% CI (−0.42 to 1.29) includes 0, so there is no evidence of additive interaction — the joint effect is consistent with independent effects. (Simulated data, so the point estimate will differ from a real study.)

#### `kk_trend_test(data, exposure_level, outcome)` / `prop_trend_test()`

Performs the Cochran-Armitage test for trend in proportions to evaluate linear trends in binomial proportions across ordinal levels.

*   **Epidemiological Example**: Evaluating if the prevalence of cardiovascular disease increases with increasing levels of daily alcohol intake (None, Low, Moderate, High).
```{r ex-05}
df_trend <- data.frame(
  alcohol = factor(rep(c("None", "Low", "Med", "High"), each = 50), levels = c("None", "Low", "Med", "High")),
  cvd = rbinom(200, 1, c(0.1, 0.15, 0.25, 0.40)[rep(1:4, each = 50)])
)
df_trend %>% kk_trend_test(alcohol, cvd)
```

> **Interpretation.** Read the *Trend OR (per unit)* row: the odds of CVD rise 1.62× (95% CI 1.17–2.24) for each step up the ordered alcohol scale (None → Low → Med → High), a significant monotonic dose-response (p = 0.003). A trend OR with a CI excluding 1 is the key evidence of increasing risk across ordinal categories.

#### `kk_nnt(data, exposure, outcome)`

Calculates the Number Needed to Treat (NNT) or Number Needed to Harm (NNH) alongside the Absolute Risk Reduction (ARR) and Relative Risk Reduction (RRR).

*   **Clinical Example**: Calculating the number of patients required to be treated with a new statin to prevent one primary cardiovascular event.
```{r ex-06}
df_treatment <- data.frame(
  statin = c(rep(1, 500), rep(0, 500)),
  event = c(rep(1, 15), rep(0, 485), rep(1, 35), rep(0, 465))
)
df_treatment %>% kk_nnt(statin, event)
```

> **Interpretation.** The statin cut the event rate from 7% to 3% — an absolute risk reduction of 4% (ARR 0.04) and a 57% relative reduction (RRR). You would need to treat 25 patients (NNT, 95% CI 15–76) to prevent one cardiovascular event; the CI excludes infinity, consistent with a significant benefit.

#### `kk_sensitivity_analysis(or_observed, p_bias)`

Computes the sensitivity of an observed Odds Ratio or Relative Risk against unmeasured confounding, assessing the strength of a potential confounder required to explain away the finding.

```{r ex-07}
# Observed OR of 2.5 with 20% estimated confounding bias prevalence
kk_sensitivity_analysis(or_observed = 2.5, p_bias = 0.20)
```

> **Interpretation.** To fully explain away the observed OR of 2.5 through unmeasured confounding, a confounder would need to be associated with the outcome by a risk ratio of at least 8.5 (given the assumed 20% prevalence difference). Such a strong hidden confounder is implausible in most settings, so the finding is relatively robust.

#### `kk_incidence_rate(data, cases, person_time, by = NULL)`

Computes incidence rates per unit of person-time with **exact (Poisson) confidence intervals**, and — when a stratifying variable is supplied — incidence **rate ratios** versus the reference level.

*   **Epidemiological Example**: Comparing the incidence rate of myocardial infarction per 1000 person-years between a high-exposure and a reference cohort.
```{r ex-08}
df_ir <- data.frame(
  arm = c("exposed", "unexposed"),
  cases = c(40, 15),
  pyears = c(1000, 1200)
)
df_ir %>% kk_incidence_rate(cases, pyears, by = arm)
```

> **Interpretation.** The exposed cohort had 40 events per 1000 person-years (exact 95% CI 28.6–54.5) versus 12.5 (7.0–20.6) in the unexposed. The rate ratio is computed against the first row (exposed = reference), so the unexposed value of 0.31 means the exposed had ~3.2× the event rate. The non-overlapping exact Poisson CIs indicate a real rate difference.

#### `kk_rr_reg(data, outcome, predictors)`

Estimates **adjusted risk ratios** for a common binary outcome using the modified Poisson approach (Zou 2004): a log-link Poisson model with robust sandwich standard errors. This is the preferred alternative to logistic regression when the outcome is frequent, since odds ratios overstate the risk ratio.

*   **Epidemiological Example**: Estimating the adjusted risk ratio of post-operative infection for an exposure, controlling for age, in a cohort where infection is common (>10%).
```{r ex-09}
df_rr <- data.frame(
  infection = rbinom(400, 1, 0.35),
  exposure = rbinom(400, 1, 0.5),
  age = rnorm(400, 50, 10)
)
kk_rr_reg(df_rr, infection, c("exposure", "age"))
```

> **Interpretation.** Each row is an adjusted risk ratio; read the `multivariable` rows for the mutually-adjusted estimates. Exposure has RR ≈ 1.01 (95% CI 0.78–1.30, p ≈ 0.96) — no association, as expected since the outcome was simulated independently. The advantage over logistic regression: these are risk ratios, directly interpretable even though infection is common.

#### `kk_rate_reg(data, outcome, predictors, person_time)` / `kk_poisson()`

Fits **Poisson rate regression** returning incidence-rate ratios (IRR), with an `offset(log(person_time))` for rate outcomes. Overdispersion is detected automatically and a **negative-binomial** model is substituted when appropriate.

*   **Epidemiological Example**: Modelling the rate of hospital admissions per person-year as a function of age and sex.
```{r ex-10}
df_rate <- data.frame(
  admissions = rpois(200, 3),
  pyears = runif(200, 50, 150),
  age = rnorm(200, 50, 10),
  sex = rbinom(200, 1, 0.5)
)
kk_rate_reg(df_rate, admissions, c("age", "sex"), person_time = pyears)
```

> **Interpretation.** Coefficients are incidence-rate ratios (per person-year via the offset). The `dispersion` column (~1.5) is checked automatically; here it stayed below the negative-binomial threshold so a Poisson fit was kept (`family = poisson`). Neither age (IRR ≈ 1.00) nor sex (IRR ≈ 1.13, p ≈ 0.14) is significant in this simulated data.

#### `kk_smd(data, treatment, variables)` / `kk_balance_table()`

Computes **standardized mean differences (SMD)** between two groups for each covariate — the standard diagnostic for baseline balance in matched, weighted, or propensity-score cohorts. Absolute SMD > 0.1 flags meaningful imbalance.

*   **Epidemiological Example**: Checking covariate balance between treated and control arms after propensity-score matching.
```{r ex-11}
df_bal <- data.frame(
  arm = rep(c("treated", "control"), each = 100),
  age = c(rnorm(100, 55, 8), rnorm(100, 50, 8)),
  smoker = rbinom(200, 1, 0.4)
)
df_bal %>% kk_smd(arm, variables = c("age", "smoker"))
```

> **Interpretation.** SMDs measure imbalance independent of sample size. Both covariates exceed the conventional |SMD| > 0.1 threshold (age −0.68, smoker −0.14) and are flagged `imbalanced = TRUE`, so the raw treated and control groups are not comparable — weighting or matching (e.g. `kk_iptw`) is warranted before estimating an effect.

#### `kk_iptw(data, treatment, outcome, covariates)`

Estimates a **propensity score**, forms **inverse-probability-of-treatment weights** (ATE or ATT, optionally stabilized), and returns the weighted **risk difference and risk ratio** with robust CIs. Use `kk_smd` on the weighted sample to confirm the covariates are balanced.

*   **Epidemiological Example**: Estimating the causal effect of a treatment on a binary outcome from observational data, adjusting for confounding by age and sex via IPTW.
```{r ex-12}
df_ip <- data.frame(
  trt = rbinom(500, 1, 0.5),
  age = rnorm(500, 50, 10),
  sex = rbinom(500, 1, 0.5)
)
df_ip$out <- rbinom(500, 1, plogis(-1 + 0.5 * df_ip$trt + 0.02 * df_ip$age))
kk_iptw(df_ip, trt, out, covariates = c("age", "sex"))
```

> **Interpretation.** After inverse-probability weighting to balance age and sex, the *average treatment effect* is a risk difference of +0.097 (95% CI 0.010–0.183) and a risk ratio of 1.18 (1.02–1.37); both exclude the null (p ≈ 0.03), so treatment raises the outcome probability by ~10 percentage points. Confirm the weights achieved balance by re-running `kk_smd` on the weighted sample.

#### `kk_epi_stats(data, exposure, outcome)`

Computes the **Odds Ratio and Relative Risk** (with confidence intervals) from two binary variables — a lightweight alternative to `kk_twobytwo` when only OR and RR are needed.

*   **Epidemiological Example**: Estimating the OR and RR of disease among exposed vs. unexposed subjects.
```{r ex-13}
df_epi <- data.frame(
  exposure = c(rep(1, 50), rep(0, 50)),
  outcome = c(rep(1, 20), rep(0, 30), rep(1, 10), rep(0, 40))
)
kk_epi_stats(df_epi, exposure, outcome)
```

> **Interpretation.** Exposed subjects had twice the risk (RR 2.0, 95% CI 1.04–3.83) and 2.67× the odds (OR 2.67) of the outcome; both CIs exclude 1, so the association is significant. As always the OR sits further from 1 than the RR, so with a 20% exposed-group risk the RR is the more faithful effect measure.

#### `odds_ratio(data, exposure, outcome)` / `risk_ratio(data, exposure, outcome)`

Convenience wrappers that return a single measure of association (the OR or the RR, respectively) with its confidence interval and p-value.

```{r ex-14}
df_or <- data.frame(exposure = c(1, 1, 0, 0, 1, 0), outcome = c(1, 0, 1, 0, 1, 0))
odds_ratio(df_or, exposure, outcome)
risk_ratio(df_or, exposure, outcome)
```

#### `kk_std_rates(data, count, pop, std_pop)`

Performs **direct age-standardization** of rates against a reference (standard) population, returning crude and standardized rates with confidence intervals.

*   **Epidemiological Example**: Age-standardizing cancer incidence to the WHO world standard population to allow comparison across regions.
```{r ex-15}
df_std <- data.frame(
  age_group = c("0-19", "20-39", "40-59", "60+"),
  cases = c(10, 25, 50, 100),
  pop = c(5000, 8000, 6000, 4000)
)
std_pop <- c(4000, 7000, 6000, 3000)
kk_std_rates(df_std, cases, pop, std_pop, multiplier = 1000)
```

> **Interpretation.** The crude rate (8.04 per 1000) is what you observe; the age-standardized rate (7.74, 95% CI 6.62–8.87) is what the population *would* experience under the reference age structure. Standardizing removes age as a confounder so regions or periods can be compared fairly. The small drop from crude to adjusted shows this population is slightly older than the standard.

#### `kk_smr(data, observed, pop, ref_rate)`

Performs **indirect standardization**, comparing observed events with those expected under a reference population's stratum-specific rates to give the **Standardized Mortality/Morbidity Ratio (SMR)** with an exact Poisson CI. The counterpart to the direct `kk_std_rates` when stratum-specific rates in the study population are unstable.

*   **Epidemiological Example**: Assessing whether mortality in an occupational cohort exceeds that expected from national age-specific death rates.
```{r ex-16}
df_smr <- data.frame(
  age_group = c("0-39", "40-59", "60+"),
  deaths = c(5, 20, 60),
  pyears = c(20000, 15000, 8000),
  ref_rate = c(0.0002, 0.0015, 0.008)
)
kk_smr(df_smr, deaths, pyears, ref_rate)
```

> **Interpretation.** 85 deaths were observed against 90.5 expected from the reference population's age-specific rates, giving an SMR of 0.94 (95% CI 0.75–1.16). Because the CI includes 1 (p = 0.60), the cohort's mortality is not significantly different from expected — no excess risk. An SMR > 1 with a CI excluding 1 would signal excess mortality.

#### `kk_risk_plot(data, title)`

Renders a **forest plot** of risk estimates (OR / RR and CIs) from the output of `kk_twobytwo`, `kk_epi_stats`, or any tibble with `Metric`, `Estimate`, `Lower`, `Upper` columns.

```{r ex-17}
res <- tibble::tibble(
  Metric = c("OR", "RR"),
  Estimate = c(1.5, 1.2),
  Lower = c(1.1, 1.0),
  Upper = c(2.0, 1.5)
)
kk_risk_plot(res)
```

---

### 2. Diagnostic Tests, Agreement & Rater Studies

#### `kk_diagnostic(data, truth, prediction)` / `confusion_metrics_ci()`

Computes key diagnostic test performance parameters: Sensitivity, Specificity, Positive/Negative Predictive Value (PPV/NPV), and Likelihood Ratios (LR+/LR-) with Wilson score confidence intervals.

*   **Clinical Example**: Evaluating the diagnostic accuracy of a rapid PCR test against the gold standard viral culture.

```{r ex-18}
df_test <- data.frame(
  gold_standard = c(rep(1, 100), rep(0, 900)),
  pcr_test = c(rep(1, 95), rep(0, 5), rep(1, 20), rep(0, 880))
)
df_test %>% kk_diagnostic(gold_standard, pcr_test)
```

> **Interpretation.** The PCR test is highly accurate: 95% sensitivity and 97.8% specificity, AUC 0.96. Note the gap between PPV (0.83) and NPV (0.99) — driven by the low 10% prevalence, a positive result is only ~83% likely to be a true case, while a negative result all but rules disease out.

#### `kk_roc(data, truth, predictor)`

Builds an ROC curve for a continuous marker, returning the **AUC with a DeLong confidence interval** and the **Youden-optimal cutoff** together with the sensitivity, specificity, PPV, and NPV achieved at that threshold.

*   **Clinical Example**: Determining the optimal biomarker cutoff and discriminative accuracy for identifying sepsis.
```{r ex-19}
df_marker <- data.frame(disease = rbinom(300, 1, 0.4))
df_marker$biomarker <- df_marker$disease * 0.8 + rnorm(300)
df_marker %>% kk_roc(disease, biomarker)
```

> **Interpretation.** The marker discriminates cases moderately well (AUC 0.73, DeLong 95% CI 0.67–0.78; 0.5 is chance). The Youden-optimal cutoff of 0.52 balances sensitivity (0.66) and specificity (0.69) — use it when you need a single decision threshold rather than the whole curve.

#### `kk_compare_roc(data, truth, predictor1, predictor2)`

Compares the AUCs of two markers measured on the same subjects using **DeLong's test** for paired ROC curves.

*   **Clinical Example**: Testing whether a novel biomarker significantly improves discrimination over an established one for the same patients.
```{r ex-20}
df_marker$established <- df_marker$disease * 0.3 + rnorm(300)
df_marker %>% kk_compare_roc(disease, biomarker, established)
```

> **Interpretation.** The first biomarker discriminates significantly better than the established one (AUC 0.73 vs 0.58; difference 0.14, DeLong p = 0.0017). Because both markers are measured on the *same* subjects, the paired DeLong test is the correct comparison — a two-sample test would ignore their correlation and overstate uncertainty.

#### `kk_calibration(data, truth, predicted)`

Assesses **calibration** of a risk model — the **Brier score**, the **Hosmer-Lemeshow** goodness-of-fit test, and a decile table of predicted vs. observed risk for a calibration plot. The complement to `kk_roc`: discrimination tells you *if* the model separates cases, calibration tells you whether the predicted probabilities are *accurate*.

*   **Clinical Example**: Checking whether a cardiovascular risk score's predicted 10-year probabilities agree with observed event rates before deploying it.
```{r ex-21}
df_cal <- data.frame(y = rbinom(500, 1, 0.3))
df_cal$p <- plogis(qlogis(0.3) + 0.8 * df_cal$y + rnorm(500))
kk_calibration(df_cal, y, p)
```

> **Interpretation.** Each row is a risk decile: compare `observed_rate` with `predicted_rate`. Good calibration means the two track closely. Here the model over-predicts in the higher-risk deciles (decile 7: 22% observed vs 44% predicted; decile 10: 56% vs 74%), so its probabilities are too high for high-risk patients. Retrieve the Brier score and Hosmer-Lemeshow test from `attr(x, "brier")` and `attr(x, "hosmer_lemeshow")`.

#### `kk_kappa(data, rater1, rater2)`

Computes Cohen's Kappa coefficient ($\kappa$), standard error, Z-statistic, and confidence intervals to evaluate inter-rater agreement for categorical classifications.

*   **Clinical Example**: Assessing diagnostic agreement between two independent neurologists classifying MRI scans as "Normal", "Inconclusive", or "Pathological".

```{r ex-22}
df_agree <- data.frame(
  neuro1 = c(rep("Normal", 50), rep("Abnormal", 50)),
  neuro2 = c(rep("Normal", 45), rep("Abnormal", 5), rep("Normal", 8), rep("Abnormal", 42))
)
df_agree %>% kk_kappa(neuro1, neuro2)
```

> **Interpretation.** Cohen's κ of 0.74 (95% CI 0.61–0.87) indicates *substantial* agreement between the two neurologists beyond chance (Landis-Koch: 0.61–0.80 = substantial). The CI excludes 0 and p is minuscule, so the agreement is well above what chance alone would produce.

#### `kk_bland_altman(data, method1, method2)`

Computes limits of agreement and constructs Bland-Altman plots to evaluate the clinical agreement between two continuous measurement methods.

*   **Clinical Example**: Assessing agreement between systolic blood pressure measurements from an arterial line vs. an oscillometric arm cuff.

```{r ex-23}
df_bp <- data.frame(
  arterial = rnorm(50, 120, 15),
  cuff = rnorm(50, 118, 14)
)
df_bp %>% kk_bland_altman(arterial, cuff)
```

#### `kk_mcnemar(data, test1, test2)`
Performs McNemar's test for paired categorical data (e.g. matched case-control designs or two diagnostic tests on the same subjects).

*   **Epidemiological Example**: Comparing the screening outcomes of Mammography and Ultrasound performed on the same cohort of patients.

```{r ex-24}
df_paired <- data.frame(
  mammography = c(rep(1, 80), rep(0, 120)),
  ultrasound = c(rep(1, 60), rep(0, 20), rep(1, 15), rep(0, 105))
)
df_paired %>% kk_mcnemar(mammography, ultrasound)
```

> **Interpretation.** McNemar's test uses only the *discordant* pairs (15 vs 20 here). The conditional OR is 1.33 with p = 0.50, so there is no significant difference in the positivity rates of the two paired tests — mammography and ultrasound detect at similar rates in this cohort.

#### `kk_confusion_matrix(x, ...)`

Computes a **full panel of classifier metrics** with confidence intervals (sensitivity, specificity, PPV, NPV, accuracy, F1, MCC, likelihood ratios, and more) directly from TP/FP/FN/TN counts. Optional bootstrap CIs via `boot = TRUE`. Supersedes the deprecated `confusion_metrics_ci()`.

*   **Clinical Example**: Full performance characterization of a screening test with 85 true positives, 10 false positives, 15 false negatives, and 890 true negatives.
```{r ex-25}
kk_confusion_matrix(c(tp = 85, fp = 10, fn = 15, tn = 890))
```

> **Interpretation.** From the four cell counts the function derives 21 metrics with exact CIs. Sensitivity is 0.85 and specificity 0.99; despite the high specificity, the 10% prevalence still shapes the predictive values (PPV 0.90, NPV 0.98). Scroll the full tibble for likelihood ratios and MCC, which summarise performance in a single prevalence-independent number.

#### `diagnostic_summary(data, truth, test)`

Produces a tidy summary of diagnostic metrics from **raw patient-level data**, applying a cutoff to numeric test results automatically.

```{r ex-26}
df_diag <- data.frame(truth = c(1, 0, 1, 0, 1, 0), test = c(1, 0, 0, 1, 1, 0))
diagnostic_summary(df_diag, truth, test)
```

#### `kk_agreement(data, rater1, rater2, weights)`

Computes **weighted Cohen's Kappa** (unweighted, linear, or quadratic weights) for ordinal rater agreement, with standard error and confidence interval.

*   **Clinical Example**: Quantifying agreement between two raters scoring tumour grade on an ordinal scale, giving partial credit for near-misses via quadratic weights.
```{r ex-27}
df_rate <- data.frame(rater1 = c(1, 0, 1, 0, 1), rater2 = c(1, 1, 0, 0, 1))
kk_agreement(df_rate, rater1, rater2, weights = "quadratic")
```

> **Interpretation.** With a binary rating the *weighted* κ is undefined (weights need ≥3 ordered categories), so read the unweighted κ (0.17) and PABAK (0.20). Both are low, but the tiny sample (n = 5) makes them unstable (CI −0.73 to 1.06). Quadratic weights pay off only when the scale is genuinely ordinal with 3+ levels, giving partial credit for near-misses.

#### `kk_icc(data, raters)`

Computes the **Intraclass Correlation Coefficient** in all six Shrout-Fleiss forms (single and average rater) with F-tests and CIs, for the agreement of continuous measurements — the continuous-scale counterpart to `kk_kappa`.

*   **Clinical Example**: Quantifying the reliability of a tumour-size measurement made by three radiologists on the same set of scans.
```{r ex-28}
ratings <- data.frame(
  r1 = c(9, 6, 8, 7, 10, 6, 8, 7, 9, 5),
  r2 = c(8, 6, 7, 8, 9, 5, 8, 6, 9, 6),
  r3 = c(9, 5, 8, 8, 9, 6, 7, 7, 8, 5)
)
kk_icc(ratings)
```

> **Interpretation.** Read `ICC2` for the reliability of a *single* rater (0.85) and `ICC2k` for the *average* of the three raters (0.94) under a two-way random model. Values above 0.75 indicate excellent reliability, so these raters are highly consistent. Pick the row matching your design: ICC1 (one-way), ICC2 (two-way random), ICC3 (two-way fixed).

#### `kk_reliability(data, items)`

Computes **Cronbach's alpha** (raw and standardized) with per-item statistics including alpha-if-item-dropped, for validating multi-item scales and questionnaires.

*   **Clinical Example**: Assessing the internal consistency of a four-item patient-reported quality-of-life scale.
```{r ex-29}
items <- data.frame(
  q1 = c(4, 5, 3, 4, 5, 2, 4, 5),
  q2 = c(4, 4, 3, 5, 5, 2, 3, 5),
  q3 = c(5, 5, 2, 4, 4, 3, 4, 4),
  q4 = c(4, 5, 3, 4, 5, 2, 4, 5)
)
kk_reliability(items)
```

> **Interpretation.** The per-item table shows each item's correlation with the scale total and the alpha the scale would have if that item were dropped. No `alpha_if_dropped` value exceeds the overall α (retrieve it via `attr(x, "alpha")` — here raw α ≈ 0.92), so every item contributes and none should be removed. α ≥ 0.7 is the usual bar for acceptable internal consistency.

#### `kk_chisq_test(data, r, c)`

Chi-square test of independence for **r × c contingency tables** (data frame or matrix input) with expected counts, residuals, and effect sizes (Cramér's V).

*   **Epidemiological Example**: Testing association between species and biting frequency across multiple categories.
```{r ex-30}
species <- c(rep("Mice", 60), rep("Gerbils", 50))
biting <- c(rep("Not", 30), rep("Mild", 30), rep("Not", 25), rep("Mild", 25))
df_chi <- data.frame(species = species, biting = biting)
kk_chisq_test(df_chi, species, biting)
```

---

### 3. Non-Parametric Tests & Baseline Group Comparisons

#### `kk_table1(data, by, variables)` / `kk_compare_groups_table()`

Builds a standard, publication-ready "Table 1" summarizing baseline demographics and clinical variables across groups, automatically performing parametric or non-parametric tests depending on variable characteristics.

*   **Clinical Example**: Summarizing age, gender, blood pressure, and smoking status at baseline stratified by placebo vs. active treatment arm.

```{r ex-31}
kk_table1(mtcars, by = "am", variables = c("mpg", "hp", "wt"))
```

> **Interpretation.** Each cell is median (IQR) by group, with an automatically chosen test in the p-value column. Automatic-transmission cars (am = 1) have higher mpg (22.8 vs 17.3, p = 0.001) and lower weight (2.32 vs 3.52, p < 0.001). This is the standard baseline-characteristics "Table 1" for a manuscript.

#### `kk_compare_groups_table(data, group, variables)` / `compare_groups_table()`

Produces a **detailed two-group comparison table** with per-group summaries, mean/proportion differences, confidence intervals, effect sizes, and p-values. Fully tidyselect- and `group_by()`-aware for stratified analyses.

*   **Clinical Example**: Comparing continuous outcomes between two treatment arms, optionally stratified by a third variable.
```{r ex-32}
# Direct comparison
kk_compare_groups_table(mtcars, am, c(mpg, hp))

# Stratified by a third variable
library(dplyr)
mtcars |>
  group_by(vs) |>
  kk_compare_groups_table(am, c(mpg, hp))
```

> **Interpretation.** Unlike Table 1, this reports the actual between-group *difference* with its CI and effect size. Manual cars average 7.24 mpg higher (p = 0.001), while the horsepower difference (−33) is not significant (p = 0.22). The stratified call repeats the comparison within each `vs` level — useful for checking effect modification.

#### `table1_summary(data, by, variables)`

A `gtsummary`-backed "Table 1" builder returning a publication-styled summary object (an alternative rendering to `kk_table1`).

```{r ex-33}
table1_summary(mtcars, by = "am", variables = c("mpg", "hp", "wt"))
```

#### `kk_median_test(data, x, group)`

Performs the Median Test for $k$ Independent Samples, comparing proportions above vs. below/equal to the composite median.

*   **Clinical Example**: Comparing the median hospital stay duration (continuous days) across three different surgical wards when stay durations are highly non-normally distributed.
```{r ex-34}
df_stay <- data.frame(
  days = c(3, 4, 2, 10, 15, 8, 4, 3, 1, 9, 12, 14, 5, 2, 7),
  ward = rep(c("Ward A", "Ward B", "Ward C"), each = 5)
)
df_stay %>% kk_median_test(days, ward)
```

> **Interpretation.** The median test dichotomises each observation at the pooled (composite) median of 5 days and compares the above/below split across wards. χ² = 0.54 on 2 df, p = 0.77 — no evidence that median stay differs by ward. It trades power for robustness, so it is most useful with heavily skewed data.

#### `kk_vdw_test(data, x, group)`

Performs the van der Waerden Normal-Scores Test as a highly powerful non-parametric alternative to ANOVA, converting ranks to normal distribution quantiles.

*   **Clinical Example**: Evaluating clinical cognitive scores under three independent noise environments when data shape assumptions are violated.

```{r ex-35}
df_noise <- data.frame(
  score = c(8, 10, 9, 10, 9, 7, 8, 5, 8, 5, 4, 8, 7, 5, 7),
  noise = rep(c("Quiet", "Classical", "Rock"), each = 5)
)
df_noise %>% kk_vdw_test(score, noise)
```

> **Interpretation.** The van der Waerden normal-scores test is a rank-based, more powerful alternative to Kruskal-Wallis/ANOVA. Here the statistic is 8.51 on 2 df, p = 0.014, so cognitive scores differ significantly across the three noise environments; the `pairwise_results` list column localises which environments differ.

---

### 4. Sequence Randomness & Serial Independence

#### `kk_runs_test(data, x, method)` / `kk_random_seq()`
Wald-Wolfowitz runs test for categorical variables, or Wallis-Moore up-down runs test for quantitative serial randomness.

*   **Epidemiological Example**: Evaluating if successive healthcare-associated infection outbreaks in an ICU follow a random temporal sequence.

```{r ex-36}
infection_seq <- c("N", "N", "Y", "N", "Y", "Y", "N", "N", "Y", "N")
kk_runs_test(infection_seq)
```

> **Interpretation.** The Wald-Wolfowitz runs test checks whether a binary sequence is randomly ordered. Observed runs (7) are close to the number expected under randomness (5.8), giving z = 0.51, p = 0.61 — no evidence of clustering or alternation, so the outbreak sequence is consistent with random timing.

#### `kk_frequency_test(data, x)`

Chi-Square Goodness-of-Fit or Binomial test checking if categories occur with equal probability.

*   **Epidemiological Example**: Verifying if congenital abnormality admissions occur uniformly across days of the week.
```{r ex-37}
admission_days <- c(1, 3, 2, 7, 5, 2, 4, 3, 1, 6, 7, 2, 5, 3)
kk_frequency_test(admission_days)
```

> **Interpretation.** A goodness-of-fit test for whether categories occur equally often. χ² = 2 on 6 df, p = 0.92 — admissions are spread evenly across the days of the week, with no evidence of a "busy day" pattern.

#### `kk_mssd_test(data, x)`

Von Neumann Mean Square Successive Difference (MSSD) test for serial correlation on continuous quantitative data.

*   **Clinical Example**: Checking if continuous blood pressure readings over time represent independent random fluctuations or exhibit serial correlation.
```{r ex-38}
bp_ticks <- c(120, 122, 121, 125, 124, 122, 120, 118, 119, 122)
kk_mssd_test(bp_ticks)
```

> **Interpretation.** The von Neumann MSSD test detects serial correlation in a continuous sequence. z = 1.68, p = 0.093 — borderline but not significant at 5%, so these blood-pressure readings are (just) consistent with independent fluctuations rather than a drifting/autocorrelated trend.

---

### 5. Regression Modeling & Proportions Comparisons

#### `kk_reg(data, outcome, predictors)` / `regression_analysis()` / `krk_reg()`
A unified modeling wrapper that automatically detects binomial outcomes (triggering Logistic Regression with odds ratios and ROC curve diagnostics) or continuous outcomes (triggering Linear Regression with check plots).

*   **Epidemiological Example**: Modeling the risk of hypertension based on age, BMI, and family history.
```{r ex-39}
regression_analysis(mtcars, outcome = "am", predictors = c("mpg", "wt"))
```

> **Interpretation.** The wrapper returns both univariate and multivariable models in one tidy frame (`model_type` column). Univariately both mpg and wt predict transmission, but in the multivariable model mpg loses significance (p = 0.99) while wt remains (coef −0.36, p = 0.015) — evidence that mpg and weight are collinear and weight carries the signal. For a binary outcome the estimates would be odds ratios with ROC diagnostics attached.

#### `compare_proportions(data)` / `compare_proportions_by()`

Pairwise comparison of proportions utilizing normal approximations with multiple comparison adjustments (e.g. Holm/Bonferroni).

```{r ex-40}
df_prop <- data.frame(
  proportion = c(0.3, 0.5, 0.25),
  trials = c(100, 120, 90),
  clinic = c("A", "B", "C")
)
compare_proportions(df_prop)
```

> **Interpretation.** All pairwise differences in proportion with unpooled (Wald) z-tests and Holm-adjusted p-values. Clinic B differs significantly from both A and C (adjusted p = 0.004 and 0.0003), while A vs C does not (adjusted p = 0.44). Read `adj_p_value` — not the raw `p_value` — when drawing conclusions across multiple comparisons.

#### `pcit(data, conf.level)`

Computes **binomial confidence intervals for proportions** from a data frame of successes and trials (the first two numeric columns), returning the proportion and its CI per row.

```{r ex-41}
df_ci <- data.frame(successes = c(10, 20), trials = c(100, 100), group = c("A", "B"))
pcit(df_ci)
```

#### `compare_proportions_kk_glm(data, group, x, n)`

Compares proportions between groups using **logistic regression with robust (sandwich) standard errors**, supporting covariate adjustment and stratification — more flexible than the normal-approximation `compare_proportions`.

*   **Epidemiological Example**: Comparing event rates across treatment arms while adjusting for a confounder.
```{r ex-42}
df_glm <- data.frame(group = c("A", "B"), x = c(15, 25), n = c(50, 50))
compare_proportions_kk_glm(df_glm, group, x, n)
```

> **Interpretation.** The estimate is the difference in proportion (A − B = −0.20) from a logistic model with robust SEs; p = 0.037, so group B's event rate is significantly higher. Unlike `compare_proportions`, this route accepts covariates and strata for adjusted contrasts.

#### `power_proportions(n, p1, p2, power, sig.level)`

Power / sample-size calculation for **two-sample proportion tests**; supply all but one parameter to solve for the missing one.

*   **Clinical Example**: Finding the power to detect a difference between a 50% and 60% response rate with 100 patients per arm.
```{r ex-43}
power_proportions(n = 100, p1 = 0.5, p2 = 0.6)
```

> **Interpretation.** With 100 patients per arm, a study comparing a 50% vs 60% response rate has only **29% power** — far below the conventional 80% target. In other words this design would miss a true 10-point difference most of the time; you would need a substantially larger sample. Leave `power` out and supply `n` to solve for power, or leave `n` out to solve for the required sample size.

#### `plot_proportion_comparisons(results)`

Renders a **forest plot** of the pairwise differences produced by `compare_proportions` or `compare_proportions_kk_glm`.

```{r ex-44}
res <- compare_proportions(df_prop)
plot_proportion_comparisons(res)
```

#### `kk_compare_independent_correlations(r, n)`

Fisher Z-transformation tests evaluating differences between correlation coefficients obtained from independent samples.

*   **Epidemiological Example**: Comparing the correlation of daily dietary sodium intake and systolic blood pressure in males vs. females.

```{r ex-45}
# Male: r = 0.65 (n=50), Female: r = 0.40 (n=60)
kk_compare_independent_correlations(c(0.65, 0.40), c(50, 60))
```

> **Interpretation.** A Fisher z-test for whether two correlations from *independent* samples differ. z = 1.78, p = 0.074 — the male (r = 0.65) and female (r = 0.40) correlations are not significantly different at 5%, despite the apparent gap, because the samples are modest. `common_r` (0.53) is the pooled estimate under the null.

#### `kk_compare_dependent_correlations(rxz, ryz, rxy, n)`

Steiger's t-test comparing two dependent correlations sharing a common criterion variable within the same sample.

*   **Epidemiological Example**: Evaluating if daily sugar intake ($X$) is a significantly stronger predictor of dental cavities ($Z$) than daily sodium intake ($Y$) in the same cohort.

```{r ex-46}
# BP vs Sugar (rxz) = 0.72, BP vs Salt (ryz) = 0.35, Sugar vs Salt (rxy) = 0.28, n = 50
kk_compare_dependent_correlations(0.72, 0.35, 0.28, 50)
```

> **Interpretation.** Steiger's test compares two correlations that share a variable *within the same sample*. Sugar correlates with cavities more strongly (0.72) than salt does (0.35), and the difference is significant (t = −2.95, df = 47, p = 0.005) — sugar is the stronger predictor even after accounting for the sugar–salt correlation (0.28).

#### `kk_firth(data, outcome, predictors)`

Fits **Firth-penalized logistic regression**, which yields finite, bias-reduced odds ratios under complete or quasi-complete **separation** (rare events or a zero cell) where ordinary logistic regression breaks down. Separation is detected and reported.

*   **Epidemiological Example**: Estimating an odds ratio for a rare adverse event where one exposure category has zero events, so standard logistic regression returns an infinite estimate.
```{r ex-47}
df_sep <- data.frame(
  y = c(0, 0, 0, 0, 1, 1, 1, 1),
  x = c(1, 2, 3, 4, 5, 6, 7, 8),
  z = rnorm(8)
)
kk_firth(df_sep, y, c("x", "z"))
```

> **Interpretation.** Here `x` perfectly separates the outcome, so ordinary logistic regression would return an infinite coefficient. Firth's penalty shrinks it to a finite, usable OR of 2.34 (95% CI 0.78–7.01). The `attr(x, "separation")` flag is `TRUE`, confirming separation was detected — the reason to use this instead of `glm()`.

---

### 6. Survival Analysis & Time-to-Event

#### `kk_survival_plot(data, time, status, group)` / `survival_plot()`
Kaplan-Meier survival curves with publication-quality formatting, confidence bands, and risk tables.
*   **Clinical Example**: Plotting time-to-death of advanced-stage lung cancer patients stratified by chemotherapy regimen.
```{r ex-48}
library(survival)
kk_survival_plot(lung, time, status, sex)
```

#### `kk_coxph(data, time, status, predictors)`

Fits univariate and multivariable **Cox proportional hazards models**, returning a tidy hazard-ratio table alongside the **Schoenfeld residual test** of the proportional-hazards assumption (per-term and global) plus concordance and AIC.

*   **Clinical Example**: Modelling time-to-death by age, sex, and ECOG performance status, while checking whether the proportional-hazards assumption holds.
```{r ex-49}
library(survival)
kk_coxph(lung, time, status, predictors = c("age", "sex", "ph.ecog"))
```

> **Interpretation.** Read the `multivariable` rows for mutually-adjusted hazard ratios: female sex (sex = 2) roughly halves the hazard of death (HR 0.58, 95% CI 0.41–0.80) and worse ECOG performance raises it (HR 1.59 per level), while age is not independently significant. Crucially, every `ph_p` (Schoenfeld test) is > 0.05, so the proportional-hazards assumption holds and the HRs are valid; had `ph_p` been small, you would switch to `kk_rmst`.

#### `kk_logrank(data, time, status, group)`

Compares survival across groups with the **log-rank test** (`rho = 0`) or the Peto-Peto / Gehan-Wilcoxon weighting (`rho = 1`), returning per-group observed/expected counts and the overall statistic.

*   **Clinical Example**: Testing whether survival differs between two chemotherapy regimens.
```{r ex-50}
library(survival)
kk_logrank(lung, time, status, sex)
```

> **Interpretation.** Group 1 (male) had more deaths than expected (observed 112 vs expected 91.6; O/E 1.22) and group 2 (female) fewer (53 vs 73.4; O/E 0.72). The log-rank test is significant (χ² = 10.3, df = 1, p = 0.0013), so survival differs by sex — consistent with the protective HR seen in `kk_coxph`.

#### `kk_rmst(data, time, status, group, tau)`

Computes the **Restricted Mean Survival Time** (area under the KM curve up to a horizon `tau`) for each group, plus the between-group **difference and ratio** with confidence intervals — an assumption-light effect measure to reach for when `kk_coxph` flags a proportional-hazards violation.

*   **Clinical Example**: Reporting the mean survival time gained (in days, within the first year) for women vs. men with lung cancer, without assuming proportional hazards.
```{r ex-51}
library(survival)
kk_rmst(lung, time, status, sex, tau = 365)
```

> **Interpretation.** Within the first year, women lived on average 297 days versus 241 for men — a **56-day gain** (95% CI 26.7–85.3, p < 0.001), equivalently an RMST ratio of 1.23. Unlike the hazard ratio, this is a direct, clinically meaningful contrast in mean survival time that needs no proportional-hazards assumption, which is why it is the fallback when `kk_coxph`'s Schoenfeld test fails.

#### `kk_cuminc(data, time, status, group, cause)`

Estimates the **competing-risks cumulative incidence** (Aalen-Johansen) for an event of interest in the presence of competing events, with **Gray's test** across groups (when the `cmprsk` package is installed).

*   **Epidemiological Example**: Estimating the cumulative incidence of relapse when death without relapse is a competing risk, compared across two treatment arms.
```{r ex-52}
df_cr <- data.frame(
  time = rexp(200, 0.1),
  status = sample(0:2, 200, replace = TRUE), # 0 = censored, 1 = relapse, 2 = death
  arm = rep(c("A", "B"), each = 100)
)
df_cr %>% kk_cuminc(time, status, arm, cause = 1)
```

> **Interpretation.** The cumulative incidence of relapse (cause 1), accounting for death as a competing risk, reaches ~0.52 in arm A and ~0.44 in arm B by the reported times. Using the Aalen-Johansen estimator (not naïve 1 − KM) avoids over-stating incidence when competing events remove people from risk. Gray's test — available via `attr(x, "gray_test")` when `cmprsk` is installed — formally compares the two curves.

---

### 7. Demographics & EGN Utilities (Bulgarian Registry)

#### `extract_egn_info(egn_vector)` / `extract_age_from_egn()`
Parses, validates, and extracts demographic profiles (Date of Birth, Gender, Age, and Birth Region) from Bulgarian Personal Identification Numbers (EGN).
*   **Epidemiological Example**: Cleaning and automatically extracting birth dates, gender, and geographic cohorts from Bulgarian electronic medical records.
```{r ex-53}
egn_sample <- c("9201014321", "8812128765")
extract_egn_info(egn_sample)
```

---

### 8. Descriptive Statistics & Summaries

#### `kk_summary(data, col)` / `comprehensive_summary()`

Computes an extensive numeric summary for a variable — central tendency, dispersion, robust estimators (Huber M), skewness/kurtosis, and normality tests — with full `group_by()` support.

*   **Clinical Example**: Describing the distribution of a biomarker overall and by treatment arm before modelling.
```{r ex-54}
kk_summary(mtcars, mpg)

# Grouped and abbreviated
library(dplyr)
mtcars |>
  group_by(am) |>
  kk_summary(mpg, verbose = "basic")
```

---

### 9. Time-Series Analysis

#### `kk_time_series(data, value_col, date_col)` / `time_series_analysis()`

Analyzes a time series end-to-end: decomposition, stationarity tests (ADF/KPSS), autocorrelation, and automatic ARIMA modelling, with optional grouping.

*   **Epidemiological Example**: Characterising a monthly surveillance count series and fitting a forecasting model.
```{r ex-55}
date <- seq(as.Date("2020-01-01"), by = "month", length.out = 24)
value <- 100 + cumsum(rnorm(24, 2, 5))
df_ts <- data.frame(date = date, value = value)
kk_time_series(df_ts)
```

#### `kk_time_metrics(data, value_col, date_col)`

Returns compact time-series descriptors — entropy, stability, linearity, and trend strength — useful for screening many series at once.
```{r ex-56}
kk_time_metrics(df_ts)
```

---

### 10. Visualization

#### `kkplot(..., rangeframe, minor_ticks, cap)`

A drop-in `ggplot()` replacement applying **Edward Tufte's data-ink principles** to the axes. **By default it behaves exactly like `ggplot()` with capped axes** (`cap = "both"`), which is safe for any `aes()` mapping and complex/faceted data. The Tufte range frame is strictly **optional**.

The Tufte options are named arguments that come *after* `...`, so always pass them **by name**:

```{r ex-57}
library(ggplot2)

# Default — capped axes, works with any data / aes() (recommended default)
kkplot(mtcars, aes(x = mpg, y = wt)) + geom_point()

# Optional: true Tufte range frame (axis lines span exactly the data range).
# Best for plots with continuous x AND y; pass the option BY NAME.
kkplot(mtcars, aes(x = mpg, y = wt), rangeframe = TRUE) + geom_point()

# Optional: subtle minor ticks for precise value reading
kkplot(mtcars, aes(x = mpg, y = wt), minor_ticks = TRUE) + geom_point()

# Combine, or change the cap style
kkplot(mtcars, aes(x = mpg, y = wt), rangeframe = TRUE, minor_ticks = TRUE) + geom_point()
kkplot(mtcars, aes(x = mpg, y = wt), cap = "none") + geom_point()
```

| Argument | Default | Effect |
|---|---|---|
| `cap` | `"both"` | Cap axis lines at the outer ticks (`"both"`, `"lower"`, `"upper"`, `"none"`). |
| `rangeframe` | `FALSE` | Draw a true Tufte range frame spanning the data range (continuous x/y only). |
| `minor_ticks` | `FALSE` | Add subtle minor ticks to both axes. |

Pairs with the Tufte theme applied by `set_plot_font()`.

#### `univariate_plot(data, variable)` / `univariate_cat_plot()` / `univariate_cont_plot()`

Automatically renders the appropriate univariate figure — bar chart for categorical variables, density/histogram for continuous — with optional grouping. The `_cat_` and `_cont_` variants (also aliased as `univariate_categorical_plot()` and `univariate_continuous_plot()`) force a specific type.
```{r ex-58}
univariate_plot(mtcars, "mpg")          # continuous -> density
univariate_plot(mtcars, "am")           # categorical -> bar
univariate_cont_plot(mtcars, "mpg", group = "cyl")
univariate_cat_plot(mtcars, "am", group = "cyl")
```

#### `kk_fullcorplot(data, method)`

Produces a full **correlation matrix plot** with significance annotations (Pearson, Spearman, or Kendall), with optional multiple-comparison adjustment.
```{r ex-59}
kk_fullcorplot(mtcars, method = "kendall")
```

#### `set_plot_font(font, size)`

Sets a global ggplot2 theme font (loaded via `sysfonts`/`showtext`) for consistent typography across all figures.
```{r ex-60, eval=FALSE}
set_plot_font("Roboto Condensed", size = 14)
```

---

### 11. Data Utilities & Setup

#### `kkonehot(data, column)` / `one_hot_encode()`

One-hot encodes a categorical column into indicator variables.
```{r ex-61}
df_oh <- tibble::tibble(id = 1:3, color = c("red", "blue", "red"))
kkonehot(df_oh, "color")
```

#### `mutate_round(data, digits)`

Rounds all numeric columns in a data frame to a given number of digits (half-up rounding).
```{r ex-62}
mutate_round(data.frame(a = c(1.234, 5.678), b = c("x", "y")), 1)
```

#### `format_tibble(data, digits)`

Formats numeric values in a results tibble for display (fixed decimals, thousands separators).
```{r ex-63}
format_tibble(tibble::tibble(Value = c(10.567, 2.3, NA)))
```

#### `kk_setup(cores, scipen)`

One-call session setup: configures parallel cores, the Bayesian backend, display options, and disables scientific notation.
```{r ex-64, eval=FALSE}
kk_setup()
```

---

## License

MIT

## Citation

If you use this package in your research, please cite it as follows:

```bibtex
@Software{kkstatfun,
  title = {kkstatfun: R Statistical Analysis Toolkit for Medical statistics and Epidemiology},
  author = {Kostadinov, Kostadin},
  year = {2026},
  version = {0.1.17},
  url = {https://github.com/kostadinoff/kkstatfun},
  doi = {10.5281/zenodo.18936020},
}
```

## Credits

This package was architected, refactored, and enhanced by **Antigravity**, an advanced AI coding assistant developed by **Google DeepMind**. 
